A Quasi-Genuine FIFO Total Order Multicast Primitive

### 2011-07-01 textos/optpaxos/draftv3.tex

The complicating factor is the possibility of a message having at least one destination group different from its source group. So, all involved groups must somehow agree regarding the delivery order of these messages. However, from assumption \emph{A1}, each group knows which other groups it could send messages to -- or receive messages from. We can use this by defining \emph{barriers} for multicast, such that \mbox{$barrier(G_{send},G_{recv}) = t$} means that the group $G_{send}$ promised that it would send no more messages with a timestamp lower than $t$ to group $G_{recv}$. We have defined that $sendersTo(G) =$ \mbox{$\{G' \neq G : G'\text{ is able to send a message to }G\}$}. When a process $p$, from group $G$, has received all the barrier values from all the groups in $sendersTo(G)$, and they are all greater than a value $t$, then $p$ knows that no more messages with timestamp lower than $t$ are coming from other groups and that, once the local ordering (the ordering of messages originated in $G$) is done, all the pending messages with timestamp up to $t$ can be delivered. Besides, a barrier is sent along with the bundle of all messages with timestamp greater than the last previous barrier sent from $G_{send}$ to $G_{recv}$, so that when a process has received a barrier from a group, it means that it knows all the messages sent by that group until the time value stored in that barrier.

@@

However, we are using the timestamp given at the creation of a message by its sender. Therefore, it might be the case that, after a message $m'$ has been proposed by a process $p$ in a consensus instance, a message $m : m.ts < m'.ts$ arrives at $p$. If $m$ is delivered with its original timestamp, the timestamp order is violated and there is no sense in using these timestamps for barriers. There are two possible solutions for that: either the message is simply discarded, and no violation to the timestamp order takes place, or we can change the timestamp of $m$ to something greater than the timestamp of $m'$. If uniform validity is to be ensured, discarding $m$ is not an option. As we need, then, to change the value of $m.ts$, two things should be noted: first, as we are using reliable FIFO channels, then the process which sent $m$ is different than that which sent $m'$, so inverting their order does not violate FIFO; finally, increasing a message timestamp must be done with caution, so that messages created by different groups at the same time have roughly the same timestamp and the barrier mechanism is efficient -- if a long sequence of messages have their timestamps increased, the last one of them may wait a long time until all the barriers required to deliver it have arrived.

@@

For every message $m$, a given group $G$ may be its source or not. Therefore, each process $p \in G$ will handle $m$ accordingly:

\begin{itemize}
  \item The message $m$ was originated in $G$:

  When a process in $G$ \cms{}s $m$, it is first \rmc{} to the other processes of $G$. Then, when $m$ is \rmd{}ed at $p$, $p$ checks whether the latest consensus instance $I_{k}$ in which it participated, or is trying to start, has already been decided -- if not, $p$ enqueues $m$ in a $propPending$ queue as the next message being proposed by it in the consensus instance $I_{k+1}$, so other tasks can keep being executed. Then, once $I_{k}$ has been decided, $p$ may start a new instance. Before that, all messages in $propPending$ that have been already decided are discarded from $propPending$. The rest is proposed as a batch in $I_{k+1}$.

  Once $m$ is decided, it is not immediately delivered to the application. Instead, $p$ checks whether some message $m_{prv} : m_{prv}.ts \geq m.ts$ has been decided previously. If that is the case, the value of $m.ts$ is changed to a value greater than the timestamp of any other message previously decided within $G$. Then, $m$ is inserted into a $barPending$ list for later being delivered, which will happen once every group $G'$ in $sendersTo(G)$ has already sent a message $barrier(G',G) = t_b$, such that $t_b \geq m.ts$. This is done because there could be a message $m'$ yet to come from another group $G'$, such that $m'.ts < m.ts$.

  Still, when a group $G$ is sending some message, it might be addressed to some other group or only to $G$. Therefore:

  \begin{itemize}
    \item If $G$ is the only destination of $m$:

    In this case, $p$ only inserts $m$ in the $barPending$ list for being delivered once all necesarry barriers have arrived.

    \item If there is some other group $G'$ as a destination of $m$:

    In this case, when $m$ is decided, $p$ \rmc{}s $m$ to every \mbox{$p' \in (m.dst \setminus \{G\})$}. When $m$ is \rmd{}ed by each $p' \in G'$, $p'$ checks whether it has ever inserted $m$ in its own $barPending$ list. If not, $p'$ inserts $m$ into $barPending$ and adjusts $barrier(G, G')$ to $m.ts$. As the sending of these messages from a group to another is done via a FIFO reliable channel, once a message $m$ from another group $G$ is \rmd{}ed, every message $m' : m'.ts < m.ts$ from $G$ has also already been \rmd{}ed.

  \end{itemize}
%   
%   In this case, when $m$ is received by $p$, $p$ checks whether the latest consensus instance $I_{k}$ in which it participated, or is trying to start, has already been decided -- if not, $p$ enqueues $m$ in a $propPending$ queue as the next message being proposed by it in the consensus instance $I_{k+1}$, so other tasks can keep being executed. Then, once $I_{k}$ has been decided, $p$ may start a new instance. Before that, all messages in $propPending$ that have been already decided are discarded from $propPending$. The rest is proposed as a batch in $I_{k+1}$. Once $m$ is decided, it is not immediately delivered to the application. Instead, $p$ checks whether some message $m_{prv} : m_{prv}.ts \geq m.ts$ has been decided previously. If that is the case, the value of $m.ts$ is changed to a value greater than the timestamp of any other message previously decided within $G$. Then, $m$ is inserted into a $barPending$ list for later being delivered, which will happen once every group $G'$ in $sendersTo(G)$ has already sent a message $barrier(G',G) = t$, such that $t > m.ts$. This is done because there could be a message $m'$ yet to come from another group $G'$, such that $m'.ts < m.ts$.
% 
%   \item When $m$ is originated in $G$, but it has at least one group other than $G$ as a destination:
%   
%   In this case, when $m$ is received, $p$ tries to initiate a consensus instance within $G$ to decide $m$. If $p$ cannot start the proposal now, $m$ is enqueued in $propPending$ for being proposed later along with other pending messages. Then, once $p$ may start a new instance, all messages in $propPending$ that have been already decided are discarded from $propPending$. The rest is proposed as a batch in the new consensus instance. Once any message $m$ is decided, $p$ checks whether some message $m_{prv} : m_{prv}.ts > m.ts$ has been decided previously. If that is the case, the value of $m.ts$ is changed to a value greater than the timestamp of any other message previously decided within $G$. Then, if $G \in m.dst$, $m$ is inserted in the $barPending$ list. Besides, when $m$ is decided, $p$ sends $m$ to every \mbox{$p' \in (m.dst \setminus \{G\})$}. When $m$ is received by each $p' \in G'$, $p'$ checks whether it has ever inserted $m$ in its own $barPending$ list. If not, $p'$ inserts $m$ into $barPending$ and adjusts $barrier(G, G')$ to $m.ts$. To ensure that, once a message $m$ is received from another group $G$, every message $m' : m'.ts<m.ts$ also from $G$ has already been received, every message is sent through a lossless FIFO channel\footnote{An ordinary TCP connection would be enough to provide such FIFO lossless channel. Here, we use FIFO reliable multicast.}.

@@

  In this case, when $m$ is received for the first time\footnote{Multiple processes may have sent $m$. To ensure integrity and order, only the first delivery is considered.} by $p$, $m$ is inserted into the $barPending$ list of $p$ and the value of $barrier(G',G)$ is set to $m.ts$.
\end{itemize}

>>>>

The complicating factor is the possibility of a message having at least one destination group different from its source group. So, all involved groups must somehow agree regarding the delivery order of these messages. However, from assumption \emph{A1}, each group knows which other groups it could send messages to -- or receive messages from. We can use this by defining \emph{barriers} for multicast, such that \mbox{$barrier(G_{send},G_{recv}) = t$} means that the group $G_{send}$ promised that it would send no more messages with a timestamp lower than or equal to $t$ to group $G_{recv}$. We have defined that $sendersTo(G) =$ \mbox{$\{G' \neq G : G'\text{ is able to send a message to }G\}$}. When a process $p$, from group $G$, has received all the barrier values from all the groups in $sendersTo(G)$, and they are all greater than a value $t$, then $p$ knows that no more messages with timestamp lower than $t$ are coming from other groups and that, once the local ordering (the ordering of messages originated in $G$) is done, all the pending messages with timestamp up to $t$ can be delivered. Besides, a barrier is sent along with the bundle of all messages with timestamp greater than the last previous barrier sent from $G_{send}$ to $G_{recv}$, so that when a process has received a barrier from a group, it means that it knows all the messages sent by that group until the time value stored in that barrier.

@@

Since we are using the timestamp given at the creation of a message by its sender, it might occur that, after a message $m'$ has been proposed by a process $p$ in some consensus instance, a message $m : m.ts < m'.ts$ arrives at $p$. If $m$ is delivered with its original timestamp, the timestamp order is violated and there is no sense in using these timestamps for barriers. There are two possible solutions for that: either the message is simply discarded, and no violation to the timestamp order takes place, or we can change the timestamp of $m$ to something greater than the timestamp of $m'$. If uniform validity is to be ensured, discarding $m$ is not an option. As we need, then, to change the value of $m.ts$, two things should be noted: first, as we are using reliable FIFO channels, then the process which sent $m$ is different than that which sent $m'$, so inverting their order does not violate FIFO; finally, increasing a message timestamp must be done with caution, so that messages created by different groups at the same time have roughly the same timestamp and the barrier mechanism is efficient -- if a long sequence of messages have their timestamps increased, the last one of them may wait a long time until all the barriers required to deliver it have arrived.

@@

The way each process $p$ of a group $G$ handles each message $m$ will depend on whether $G$ is the source group of $m$:

\begin{itemize}
  \item If $G$ is the source of $m$:

  When some process in $G$ \cms{}s $m$, it is first \rmc{} to the other processes of $G$. Then, when $m$ is \rmd{}ed at a process $p \in G$, $p$ checks whether the latest consensus instance $I_{k}$ in which it participated, or is trying to start, has already been decided -- if not, $p$ enqueues $m$ in a $propPending$ queue as the next message to be proposed by it in the consensus instance $I_{k+1}$, so other tasks can keep being executed. Then, once $I_{k}$ has been decided, $p$ may start a new instance. Before that, all messages in $propPending$ that have been already decided are discarded from $propPending$. The rest is proposed as a batch in $I_{k+1}$.

  Once $m$ has been decided, it is not immediately delivered to the application. Instead, $p$ checks whether some message $m' : m'.ts \geq m.ts$ has been decided previously. If that is the case, the value of $m.ts$ is changed to a value greater than the timestamp of any other message previously decided within $G$. Then, $m$ is inserted into a $barPending$ list for later being delivered, which will happen once every group $G'$ in $sendersTo(G)$ has already sent a message $barrier(G',G) = t_b$, such that $t_b \geq m.ts$. Only after that will $p$ know for sure that no message $m'' : m''.ts < m.ts$ will arrive from another group.

  After learning that $m$ has been decided, $p$ checks whether there is some other group which is also a destination of $m$. If that is the case, $p$ \rmc{}s $m$ to every \mbox{$p' \in (m.dst \setminus \{G\})$}.

%   
%   In this case, when $m$ is received by $p$, $p$ checks whether the latest consensus instance $I_{k}$ in which it participated, or is trying to start, has already been decided -- if not, $p$ enqueues $m$ in a $propPending$ queue as the next message being proposed by it in the consensus instance $I_{k+1}$, so other tasks can keep being executed. Then, once $I_{k}$ has been decided, $p$ may start a new instance. Before that, all messages in $propPending$ that have been already decided are discarded from $propPending$. The rest is proposed as a batch in $I_{k+1}$. Once $m$ is decided, it is not immediately delivered to the application. Instead, $p$ checks whether some message $m_{prv} : m_{prv}.ts \geq m.ts$ has been decided previously. If that is the case, the value of $m.ts$ is changed to a value greater than the timestamp of any other message previously decided within $G$. Then, $m$ is inserted into a $barPending$ list for later being delivered, which will happen once every group $G'$ in $sendersTo(G)$ has already sent a message $barrier(G',G) = t$, such that $t > m.ts$. This is done because there could be a message $m'$ yet to come from another group $G'$, such that $m'.ts < m.ts$.
% 
%   \item When $m$ is originated in $G$, but it has at least one group other than $G$ as a destination:
%   
%   In this case, when $m$ is received, $p$ tries to initiate a consensus instance within $G$ to decide $m$. If $p$ cannot start the proposal now, $m$ is enqueued in $propPending$ for being proposed later along with other pending messages. Then, once $p$ may start a new instance, all messages in $propPending$ that have been already decided are discarded from $propPending$. The rest is proposed as a batch in the new consensus instance. Once any message $m$ is decided, $p$ checks whether some message $m_{prv} : m_{prv}.ts > m.ts$ has been decided previously. If that is the case, the value of $m.ts$ is changed to a value greater than the timestamp of any other message previously decided within $G$. Then, if $G \in m.dst$, $m$ is inserted in the $barPending$ list. Besides, when $m$ is decided, $p$ sends $m$ to every \mbox{$p' \in (m.dst \setminus \{G\})$}. When $m$ is received by each $p' \in G'$, $p'$ checks whether it has ever inserted $m$ in its own $barPending$ list. If not, $p'$ inserts $m$ into $barPending$ and adjusts $barrier(G, G')$ to $m.ts$. To ensure that, once a message $m$ is received from another group $G$, every message $m' : m'.ts<m.ts$ also from $G$ has already been received, every message is sent through a lossless FIFO channel\footnote{An ordinary TCP connection would be enough to provide such FIFO lossless channel. Here, we use FIFO reliable multicast.}.

@@

  Although every process in $G'$ sends this message, it is done via a FIFO reliable channel, so even if messages are \rmd{}ed twice at $p$, once $m$ is \rmd{}ed, every message $m' : m'.ts < m.ts$ from $G$ has also already been \rmd{}ed. As there may be multiple \rmd{}ies of $m$, only the first one is considered, upon which $m$ is inserted into the $barPending$ list of $p$. Besides, the value of $barrier(G',G)$ is set to $m.ts$, for $p$ knows that no more messages with lower timestamps will arrive from $G'$.

\end{itemize}
